---
title: "breaking_down_targets"
output: html_document
date: "2025-02-24"
---

```{r setup, include=FALSE}
#install.packages("tidycensus")
#install.packages("censusapi")
library(tidycensus)
library(tidyverse)
library(censusapi)
```

Goals:

- Increase Vermont’s population to 802,000 residents by 2035 by recruiting and 
retaining working-age people.
  - Current population: 647,464
-  Increase labor force participation rate to 70% by 2035   
  - Add an average of 13,500 people to its workforce annually over 
    the next 10-15 years
- Increase Vermont’s non-seasonal housing stock to 350,000 homes by 2035
  - 36,000 primary homes needed in VT over the next 5 years 
    - 7,200 new homes annually over the next five years

```{r}
Sys.setenv(CENSUS_KEY = "d2c6932eca5b04592aaa4b32840c534b274382dc")

state_fips <- 50

test <- getCensus(
  name = "acs/acs5",
  vintage = 2020,
  vars = c("NAME", "B01001_001E", "B19013_001E"),
  region = "county:*",
  regionin = paste0("state:", state_fips)
)
```

### STATEWIDE AGE BREAKDOWN
```{r}
# pull data for age distributions

vt_age <- getCensus(
  name = "acs/acs5",
  vintage = 2023,
  vars = paste0("B01001_", str_pad(3:49, 3, pad="0"), "E"),
  region = "state:50"
  ) %>%
  select(-c("state"))
vt_age
```





```{r}
# pull data for age distributions

age_distribution_data <- function(year) {
  age <- getCensus(
    name = "acs/acs5",
    vintage = year,
    vars = paste0("B01001_", str_pad(3:49, 3, pad="0"), "E"),
    region = "state:50"
  ) %>%
    select(-c("state"))
  
  age_long <- as.data.frame(t(age))
  colnames(age_long) <- "population"
  age_long$variable <- rownames(age_long) 
  age_long$population <- as.numeric(age_long$population)  

  # Age group pattern mapping
  age_group_patterns <- list(
    "Under 5 years" = "B01001_0(03|27)E",  
    "5 to 9 years" = "B01001_0(04|28)E",  
    "10 to 14 years" = "B01001_0(05|29)E",  
    "15 to 19 years" = "B01001_0(06|07|30|31)E",  
    "20 to 24 years" = "B01001_0(08|09|10|32|33|34)E",  
    "25 to 29 years" = "B01001_0(11|35)E",  
    "30 to 34 years" = "B01001_0(12|36)E",  
    "35 to 39 years" = "B01001_0(13|37)E",  
    "40 to 44 years" = "B01001_0(14|38)E",  
    "45 to 49 years" = "B01001_0(15|39)E",  
    "50 to 54 years" = "B01001_0(16|40)E",  
    "55 to 59 years" = "B01001_0(17|41)E",  
    "60 to 64 years" = "B01001_0(18|19|42|43)E",  
    "65 to 69 years" = "B01001_0(20|21|44|45)E",  
    "70 to 74 years" = "B01001_0(22|46)E",  
    "75 to 79 years" = "B01001_0(23|47)E",  
    "80 to 84 years" = "B01001_0(24|48)E",  
    "85 years and over" = "B01001_0(25|49)E"
  )

  age_long$age_group <- sapply(age_long$variable, function(var) {
    matched_group <- names(Filter(function(pattern) str_detect(var, pattern), age_group_patterns))
    if (length(matched_group) > 0) return(matched_group) else return(NA)
  })

  age_summary <- age_long %>%
    group_by(age_group) %>%
    summarise(total_population = sum(population, na.rm = TRUE)) %>%
    filter(!is.na(age_group))  # Remove any unmatched rows

  return(age_summary)
}
```

```{r}
age_distribution_data(2023)
```

```{r}
  natl_census_data_raw <- getCensus(
    name = "acs/acs5",
    vintage = 2023,
    vars = c("NAME", census_variables$code, paste0("B01001_", str_pad(3:49, 3, pad="0"), "E")),
    region = "us:1"
  )
  
  natl_census_data <- natl_census_data_raw |> 
    rename_with(~ census_variables$title, .cols = any_of(census_variables$code))
```


```{r}
head(natl_census_data)
```

```{r}
plot_age_distribution <- function(state_df, national_df) {
  # Shared age bucket and label formatting
  process_age_data <- function(df) {
    df %>%
      mutate(
        age_bucket = case_when(
          age_group %in% c("Under 5 years", "5 to 9 years", "10 to 14 years") ~ "0-14 years",
          age_group %in% c("15 to 19 years", "20 to 24 years") ~ "15-24 years",
          age_group %in% c("25 to 29 years","30 to 34 years","35 to 39 years",
                           "40 to 44 years","45 to 49 years") ~ "25-49 years",
          age_group %in% c("50 to 54 years", "55 to 59 years", "60 to 64 years") ~ "50-64 years",
          age_group %in% c("65 to 69 years", "70 to 74 years", "75 to 79 years",
                           "80 to 84 years", "85 years and over") ~ "65+ years",
          TRUE ~ "Unknown"
        ),
        age_label = case_when(
          age_group == "Under 5 years" ~ "Under 5",
          age_group == "85 years and over" ~ "85+",
          TRUE ~ gsub(" to ", " - ", gsub(" years", "", age_group))
        )
      )
  }
  state_df <- process_age_data(state_df)
  national_df <- process_age_data(national_df)

  age_levels <- rev(c("85+", "80 - 84", "75 - 79", "70 - 74",
                      "65 - 69", "60 - 64", "55 - 59", "50 - 54",
                      "45 - 49", "40 - 44", "35 - 39", "30 - 34", 
                      "25 - 29", "20 - 24", "15 - 19", "10 - 14", 
                      "5 - 9", "Under 5"))
  state_df$age_label <- factor(state_df$age_label, levels = age_levels)
  national_df$age_label <- factor(national_df$age_label, levels = age_levels)

  scale_factor <- sum(state_df$total_population) / sum(national_df$total_population)

  national_density <- national_df %>%
    mutate(age_numeric = as.numeric(age_label),
           scaled_population = total_population * scale_factor) %>%
    complete(age_label = age_levels, fill = list(`Total Population` = 0)) %>%
    arrange(age_label)

  # Scale national density to match the range of state data
 # national_density$scaled_density <- scales::rescale(national_density$total_population, to = c(0, max_state))

  # Plot
  ggplot(state_df, aes(x = age_label, y = total_population, fill = age_bucket)) +
    geom_bar(stat = "identity") +
    geom_line(data = national_density, aes(x = age_label, y = scaled_population, group = 1),
              color = "black", size = 1.2) +
    labs(
      x = "Age Group", 
      y = "\nTotal Population", 
      title = "Vermont Age Group Distribution (2023)"
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_manual(values = c(
      "0-14 years" = "deepskyblue1", 
      "15-24 years" = "aquamarine2", 
      "25-49 years" = "aquamarine4", 
      "50-64 years" = "orange", 
      "65+ years" = "red"
    )) +
    theme_minimal() +
    theme(
      legend.position = "none", 
      text = element_text(family = "Georgia"),
      axis.title.x = element_text(size = 16, face = "bold"),
      axis.title.y = element_text(size = 16, face = "bold"),
      axis.text.x = element_text(size = 14, face = "bold", angle = 45, hjust = 1),
      axis.text.y = element_text(size = 14, face = "bold"),
      plot.title = element_text(size = 22, face = "bold")
    )
}
```


```{r}
build_state_age_df <- function(df) {
  # Filter for state level age demographic data
  age <- df %>%
    select(starts_with("B01001_"))  %>%
    summarise(across(everything(), ~ sum(.x, na.rm = TRUE)))
  
  age_long <- as.data.frame(t(age))
  colnames(age_long) <- "population"
  age_long$variable <- rownames(age_long) 
  age_long$population <- as.numeric(age_long$population)  
  
  # Age group pattern mapping
  age_group_patterns <- list(
    "Under 5 years" = "B01001_0(03|27)E",  
    "5 to 9 years" = "B01001_0(04|28)E",  
    "10 to 14 years" = "B01001_0(05|29)E",  
    "15 to 19 years" = "B01001_0(06|07|30|31)E",  
    "20 to 24 years" = "B01001_0(08|09|10|32|33|34)E",  
    "25 to 29 years" = "B01001_0(11|35)E",  
    "30 to 34 years" = "B01001_0(12|36)E",  
    "35 to 39 years" = "B01001_0(13|37)E",  
    "40 to 44 years" = "B01001_0(14|38)E",  
    "45 to 49 years" = "B01001_0(15|39)E",  
    "50 to 54 years" = "B01001_0(16|40)E",  
    "55 to 59 years" = "B01001_0(17|41)E",  
    "60 to 64 years" = "B01001_0(18|19|42|43)E",  
    "65 to 69 years" = "B01001_0(20|21|44|45)E",  
    "70 to 74 years" = "B01001_0(22|46)E",  
    "75 to 79 years" = "B01001_0(23|47)E",  
    "80 to 84 years" = "B01001_0(24|48)E",  
    "85 years and over" = "B01001_0(25|49)E"
  )
  
  age_long$age_group <- sapply(age_long$variable, function(var) {
    matched_group <- names(Filter(function(pattern) str_detect(var, pattern), age_group_patterns))
    if (length(matched_group) > 0) return(matched_group) else return(NA)
  })
  
  age_summary <- age_long %>%
    group_by(age_group) %>%
    summarise(total_population = sum(population, na.rm = TRUE)) %>%
    filter(!is.na(age_group)) 
  return(age_summary)
}
```


```{r}
age_groups2023 <- age_distribution_data(2023)
natl_data_age <- build_state_age_df(natl_census_data)
plot_age_distribution(age_groups2023, natl_data_age)
```
















### POPULATION STUFF 
```{r}
vt_pop <- sum(test$B01001_001E)
vt_income <- sum(test$B19013_001E)

test <- test %>%
  rename( "population" = "B01001_001E", "median_household_income" = "B19013_001E") %>%
  mutate(pop_proportion = (population/vt_pop),
         income_proportion = (median_household_income/vt_income))
head(test)
```

```{r}
pop_goal <- 802000-647464
goal_breakdown <- test %>%
  select("NAME", "income_proportion", "pop_proportion") %>%
  mutate(new_proportion_by_pop = floor(pop_proportion*pop_goal),
         new_proportion_by_income = floor(income_proportion*pop_goal),
         pop_income_index = (pop_proportion + income_proportion)/2,
         new_proportion_by_index = floor(pop_income_index_norm*pop_goal))
head(goal_breakdown)
```



```{r}
# install.packages("treemap")
# library(treemap)
```


```{r}
goal_breakdown$NAME <- trimws(sapply(strsplit(goal_breakdown$NAME, ","), `[`, 1))
goal_breakdown <- goal_breakdown %>%
  mutate(label = paste(NAME, new_proportion_by_pop, sep=": "))

treemap(goal_breakdown,
            index="label",
            vSize="new_proportion_by_pop",
            type="index", 
            fontsize.labels = c(8, 12),
            title = 
          "Breakdown of New Population in Each County by Current Population"
            )
```

```{r}
goal_breakdown$NAME <- trimws(sapply(strsplit(goal_breakdown$NAME, ","), `[`, 1))
goal_breakdown <- goal_breakdown %>%
  mutate(inc_label = paste(NAME, new_proportion_by_income, sep=": "))

treemap(goal_breakdown,
            index="inc_label",
            vSize="new_proportion_by_income",
            type="index", 
            fontsize.labels = c(8, 12),
            title = "Breakdown of New Population in Each County by Income",
            )
```

```{r}
goal_breakdown$NAME <- trimws(sapply(strsplit(goal_breakdown$NAME, ","), `[`, 1))
goal_breakdown <- goal_breakdown %>%
  # mutate(index_label = paste(NAME, "\n" , "\n", round(new_proportion_by_index/pop_goal*100,1),"%"))
  mutate(index_label = paste(NAME, "\n" , "\n", round(new_proportion_by_index, -2), "people"))
treemap(goal_breakdown,
            index="index_label",
            vSize="new_proportion_by_index",
            type="index", 
            fontsize.labels = c(8, 12),
            title = "Breakdown of New Population in Each County by Income/Population",
            )
```

```{r}
Sys.setenv(CENSUS_KEY = "d2c6932eca5b04592aaa4b32840c534b274382dc")

state_fips <- 50

towns <- getCensus(
  name = "acs/acs5",
  vintage = 2020,
  vars = c("NAME", "B01001_001E", "B19013_001E"),
  region = "county subdivision:*",
  regionin = paste0("state:", state_fips)
)

towns <- towns %>%
  filter(if_all(everything(), ~ . >= 0))
```

```{r}
drop_na(towns)
towns <- towns %>%
  rename( "population" = "B01001_001E", "median_household_income" = "B19013_001E") %>%
  separate(NAME, into = c("town", "county", "state"), sep = ", ") %>%
  mutate(
    town = str_remove_all(town, "(?i)\\btown\\b"),
    county = str_remove_all(county, "(?i)\\bcounty\\b"),
    town = str_trim(town),
    county = str_trim(county),
  )

head(towns)
```

```{r}
towns <- towns %>%
  mutate(
pop_proportion = (population/vt_pop),
    income_proportion = (median_household_income/vt_income),
    new_proportion_by_pop = floor(pop_proportion*pop_goal),
    new_proportion_by_income = floor(income_proportion*pop_goal),
    pop_income_index = (pop_proportion + income_proportion)/2,
    new_proportion_by_index = floor(pop_income_index*pop_goal)
)
head(towns)
```


```{r}
# function to make a tree map for each town in the county when clicked

town_treemap <- function(county, df){
  town_df <- df %>%
  filter(county == !!county) %>%
  mutate(label = paste(town,":", new_proportion_by_pop, "people"))
  print(summary(town_df$new_proportion_by_index))
  treemap(town_df,
            index="label",
            vSize="new_proportion_by_pop",
            type="index", 
            fontsize.labels = c(8, 12),
            title = "Breakdown of New Population in Each Town by Current Population",
            vColor = "new_proportion_by_pop", 
            palette = "Oranges"
            )
}
```

```{r}
town_treemap("Addison", towns)
```

```{r}
latent_cap <- read_csv("latent-capacity.csv")
head(latent_cap)
```

```{r}
latent_cap <- latent_cap %>%
  rename(latent_cap = `Latent Capacity`) %>%
  group_by(County) %>%
  summarise(latent_cap = sum(latent_cap))
latent_cap
```

```{r}
library(data.table)
jobs_homes <- read.csv("JobsHomesMap_data.csv", sep = "\t", fileEncoding = "UTF-16")

jobs_homes <- jobs_homes %>%
  drop_na() %>%
  rename(jobs_homes_index = Jobs.Homes.Index) %>%
  group_by(County) %>%
  summarise(jobs_homes_index = mean(jobs_homes_index)) %>%
  mutate(County = str_trim(str_remove(County, "County")))
jobs_homes
```

```{r}
library(readxl)
school_latency <- read_excel("school_latency.xlsx", sheet = "Data")
school_latency <- school_latency %>%
  rename(latent_cap = `Latent Capacity`) %>%
  drop_na() %>%
  group_by(County) %>%
  summarise(latent_cap_school = sum(latent_cap)) 
school_latency
```

```{r} 
county_caps <- left_join(latent_cap, jobs_homes )
county_caps <- left_join(county_caps, school_latency)
county_caps
```
```{r}
test <- county_caps %>% filter(County == "Addison")
test_long <- test %>%
  pivot_longer(cols = -County, names_to = "Metric", values_to = "Value")

# Create the plot
ggplot(test_long, aes(x = Metric, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge")  + 
  labs(title = "County-Level Metrics", x = "Metric", y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
  pop <- getCensus(
    name = "acs/acs5",
    vintage = 2020,
    vars = c("NAME", "B01001_001E"),
    region = "county:*",
    regionin = paste0("state:", state_fips)
  )  
vt_pop <- sum(pop$B01001_001E)

   pop <- pop %>% 
     rename( "population" = "B01001_001E") %>%
     mutate(pop_goal = floor((population/vt_pop)*(802000 - 647464)),
            County = str_trim(str_remove(NAME, "County, Vermont"))
           ) %>% select(County, pop_goal)
   head(pop)
```


```{r}
## using 18:1
library(readxl)
pth <- getwd()
school <- read_excel(paste0(pth, "/data/teacher_information.xlsx")) %>%
  mutate(latency = (num_teachers*18)-(num_teachers*student_teacher_ratio)) %>%
  distinct(County, school_district, latency) %>% 
  group_by(County) %>%
  summarise(latency = sum(latency)) 
head(school, 40)
```
```{r}
library(sf)
addison <- read_sf("ACRPC_Zoning.geojson")
unique(addison$Jurisdiction)
```

```{r}
head(addison)

addison |>
  filter(str_detect(`Jurisdiction District Name`, "Middlebury")) |>
  # ggplot() +
  # geom_sf(aes(fill = `4F Allowance`))
  leaflet() |>
  addTiles() |>
  addPolygons(label = ~`Jurisdiction District Name`,
                fill = ~`1F Allowance`)
```


















