---
title: "breaking_down_targets"
output: html_document
date: "2025-02-24"
---

```{r setup, include=FALSE}
#install.packages("tidycensus")
#install.packages("censusapi")
library(tidycensus)
library(tidyverse)
library(censusapi)
```

Goals:

- Increase Vermont’s population to 802,000 residents by 2035 by recruiting and 
retaining working-age people.
  - Current population: 647,464
-  Increase labor force participation rate to 70% by 2035   
  - Add an average of 13,500 people to its workforce annually over 
    the next 10-15 years
- Increase Vermont’s non-seasonal housing stock to 350,000 homes by 2035
  - 36,000 primary homes needed in VT over the next 5 years 
    - 7,200 new homes annually over the next five years

```{r}
Sys.setenv(CENSUS_KEY = "d2c6932eca5b04592aaa4b32840c534b274382dc")

state_fips <- 50

test <- getCensus(
  name = "acs/acs5",
  vintage = 2020,
  vars = c("NAME", "B01001_001E", "B19013_001E"),
  region = "county:*",
  regionin = paste0("state:", state_fips)
)
```

### STATEWIDE AGE BREAKDOWN
```{r}
# pull data for age distributions

vt_age <- getCensus(
  name = "acs/acs5",
  vintage = 2023,
  vars = paste0("B01001_", str_pad(3:49, 3, pad="0"), "E"),
  region = "state:50"
  ) %>%
  select(-c("state"))
```


```{r}
# pull data for age distributions

age_distribution_data <- function(year) {
  # I will later add functionality to filter by county so we can
  # get the age distribution by county in the app 
  age <- getCensus(
    name = "acs/acs5",
    vintage = year,
    vars = paste0("B01001_", str_pad(3:49, 3, pad="0"), "E"),
    region = "state:50"
  ) %>%
    select(-c("state"))
  
  age_long <- as.data.frame(t(age))
  
  age_group_patterns <- list(
    "Under 5 years" = "B01001_00(3|27)E",  
    "5 to 9 years" = "B01001_00(4|28)E",  
    "10 to 14 years" = "B01001_00(5|29)E",  
    "15 to 19 years" = "B01001_00(6|7|30|31)E",  
    "20 to 24 years" = "B01001_00(8|9|10|32|33|34)E",  
    "25 to 29 years" = "B01001_00(11|35)E",  
    "30 to 34 years" = "B01001_00(12|36)E",  
    "35 to 39 years" = "B01001_00(13|37)E",  
    "40 to 44 years" = "B01001_00(14|38)E",  
    "45 to 49 years" = "B01001_00(15|39)E",  
    "50 to 54 years" = "B01001_00(16|40)E",  
    "55 to 59 years" = "B01001_00(17|41)E",  
    "60 to 64 years" = "B01001_00(18|19|42|43)E",  
    "65 to 69 years" = "B01001_00(20|21|44|45)E",  
    "70 to 74 years" = "B01001_00(22|46)E",  
    "75 to 79 years" = "B01001_00(23|47)E",  
    "80 to 84 years" = "B01001_00(24|48)E",  
    "85 years and over" = "B01001_00(25|49)E"
  )

  
  
  age_long <- as.data.frame(age_long)
  age_long$code <- rownames(age_long)
  colnames(age_long) <- c("population", "code") 
  
  age_groups <- data.frame(
    age_group = names(age_group_mapping),
    total_population = sapply(age_group_mapping, function(cols) {
      sum(age_long$population[age_long$code %in% cols], na.rm = TRUE)
    })
  ) %>% select("total_population") %>%
    rownames_to_column(var = "age_group")
  
  return(age_groups)
}
```

```{r}
age_distribution_data(2023)
```







```{r}
# Create a new column for grouping into buckets
plot_age_distribution <- function(df) {
  age_groups <- df %>%
  mutate(age_bucket = case_when(
    age_group %in% c("Under 5 years", "5 to 9 years", "10 to 14 years") ~ "0-14 years",
    age_group %in% c("15 to 19 years", "20 to 24 years") ~ "15-24 years",
    age_group %in% c("25 to 29 years","30 to 34 years","35 to 39 years","40 to 44 years","45 to 49 years") ~ "25-49 years",
    age_group %in% c("50 to 54 years", "55 to 59 years", "60 to 64 years") ~ "50-64 years",
    age_group %in% c("65 to 69 years", "70 to 74 years", "75 to 79 years", "80 to 84 years", "85 years and over") ~ "65+ years",
    TRUE ~ "Unknown"
  ))

# Convert age_group to a factor with correct ordering
age_groups$age_group <- factor(age_groups$age_group, levels = c(
  "85 years and over", "80 to 84 years", "75 to 79 years", "70 to 74 years",
"65 to 69 years", "60 to 64 years", "55 to 59 years", "50 to 54 years",
"45 to 49 years", "40 to 44 years", "35 to 39 years",
"30 to 34 years", "25 to 29 years","20 to 24 years","15 to 19 years","10 to 14 years", 
"5 to 9 years", "Under 5 years"
))

# Create the bar chart with color-coded buckets
ggplot(age_groups, aes(y = age_group, x = total_population, fill = age_bucket)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = total_population), vjust = 0.3, hjust = 3, color="white", size = 3) +
  labs(x = "Total Population", y = "Age Group", title = "Vermont Age Group Distribution from 2023") +
  theme_minimal() +
  scale_fill_manual(values = c("0-14 years" = "deepskyblue1", "15-24 years" = "aquamarine2", 
                               "25-49 years" = "aquamarine4", "50-64 years" = "orange", 
                               "65+ years" = "red")) +
  theme(legend.position = "none")
}
```


```{r}
age_groups2023 <- age_distribution_data(2023)
plot_age_distribution(age_groups2023)
```
















### POPULATION STUFF 
```{r}
vt_pop <- sum(test$B01001_001E)
vt_income <- sum(test$B19013_001E)

test <- test %>%
  rename( "population" = "B01001_001E", "median_household_income" = "B19013_001E") %>%
  mutate(pop_proportion = (population/vt_pop),
         income_proportion = (median_household_income/vt_income))
head(test)
```

```{r}
pop_goal <- 802000-647464
goal_breakdown <- test %>%
  select("NAME", "income_proportion", "pop_proportion") %>%
  mutate(new_proportion_by_pop = floor(pop_proportion*pop_goal),
         new_proportion_by_income = floor(income_proportion*pop_goal),
         pop_income_index = (pop_proportion + income_proportion)/2,
         new_proportion_by_index = floor(pop_income_index_norm*pop_goal))
head(goal_breakdown)
```



```{r}
# install.packages("treemap")
# library(treemap)
```


```{r}
goal_breakdown$NAME <- trimws(sapply(strsplit(goal_breakdown$NAME, ","), `[`, 1))
goal_breakdown <- goal_breakdown %>%
  mutate(label = paste(NAME, new_proportion_by_pop, sep=": "))

treemap(goal_breakdown,
            index="label",
            vSize="new_proportion_by_pop",
            type="index", 
            fontsize.labels = c(8, 12),
            title = 
          "Breakdown of New Population in Each County by Current Population"
            )
```

```{r}
goal_breakdown$NAME <- trimws(sapply(strsplit(goal_breakdown$NAME, ","), `[`, 1))
goal_breakdown <- goal_breakdown %>%
  mutate(inc_label = paste(NAME, new_proportion_by_income, sep=": "))

treemap(goal_breakdown,
            index="inc_label",
            vSize="new_proportion_by_income",
            type="index", 
            fontsize.labels = c(8, 12),
            title = "Breakdown of New Population in Each County by Income",
            )
```

```{r}
goal_breakdown$NAME <- trimws(sapply(strsplit(goal_breakdown$NAME, ","), `[`, 1))
goal_breakdown <- goal_breakdown %>%
  # mutate(index_label = paste(NAME, "\n" , "\n", round(new_proportion_by_index/pop_goal*100,1),"%"))
  mutate(index_label = paste(NAME, "\n" , "\n", round(new_proportion_by_index, -2), "people"))
treemap(goal_breakdown,
            index="index_label",
            vSize="new_proportion_by_index",
            type="index", 
            fontsize.labels = c(8, 12),
            title = "Breakdown of New Population in Each County by Income/Population",
            )
```

```{r}
Sys.setenv(CENSUS_KEY = "d2c6932eca5b04592aaa4b32840c534b274382dc")

state_fips <- 50

towns <- getCensus(
  name = "acs/acs5",
  vintage = 2020,
  vars = c("NAME", "B01001_001E", "B19013_001E"),
  region = "county subdivision:*",
  regionin = paste0("state:", state_fips)
)

towns <- towns %>%
  filter(if_all(everything(), ~ . >= 0))
```

```{r}
drop_na(towns)
towns <- towns %>%
  rename( "population" = "B01001_001E", "median_household_income" = "B19013_001E") %>%
  separate(NAME, into = c("town", "county", "state"), sep = ", ") %>%
  mutate(
    town = str_remove_all(town, "(?i)\\btown\\b"),
    county = str_remove_all(county, "(?i)\\bcounty\\b"),
    town = str_trim(town),
    county = str_trim(county),
  )

head(towns)
```

```{r}
towns <- towns %>%
  mutate(
pop_proportion = (population/vt_pop),
    income_proportion = (median_household_income/vt_income),
    new_proportion_by_pop = floor(pop_proportion*pop_goal),
    new_proportion_by_income = floor(income_proportion*pop_goal),
    pop_income_index = (pop_proportion + income_proportion)/2,
    new_proportion_by_index = floor(pop_income_index*pop_goal)
)
head(towns)
```


```{r}
# function to make a tree map for each town in the county when clicked

town_treemap <- function(county, df){
  town_df <- df %>%
  filter(county == !!county) %>%
  mutate(label = paste(town,":", new_proportion_by_pop, "people"))
  print(summary(town_df$new_proportion_by_index))
  treemap(town_df,
            index="label",
            vSize="new_proportion_by_pop",
            type="index", 
            fontsize.labels = c(8, 12),
            title = "Breakdown of New Population in Each Town by Current Population",
            vColor = "new_proportion_by_pop", 
            palette = "Oranges"
            )
}
```

```{r}
town_treemap("Addison", towns)
```